---
title: "Model version 2024-10-11"
output: html_document
date: "2024-10-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Load Packages

```{r load-packages, echo=FALSE}
# Load required R packages for data manipulation, statistical distributions, and parallel processing
library(tidyverse)      # Data manipulation: filter, mutate, select, group_by, left_join, etc.
library(fitdistrplus)   # Fit probability distributions to data
library(rmutil)         # Laplace distribution functions
library(truncnorm)      # Truncated normal distribution functions
library(EnvStats)       # Environmental statistics: truncated normal and triangular distributions
library(MCMCpack)       # Dirichlet distribution for sampling relative abundances
library(furrr)          # Parallel processing with future_map() and plan()
library(deSolve)        # Ordinary differential equation solver (used by predict_dynamic_growth)
```

## 2. Load Functions and Input Files

```{r load-functions-data, echo=FALSE}
# Source custom utility functions for growth modeling
source("UtilityFunctions_dynamic_growth.R")
source("UtilityFunctions.R")

# Load input data files
# Initial concentration data from bulk tank
Log10InitConc <- read.csv("Count initial data.csv", sep=";")

# Categorization data for relative abundance, log reduction, and growth potential
genus_rel_ab_cat <- read.csv("Abundance bins 20250501.csv", sep =";")
bin_alphas <- read.csv("Bin Alphas (1, 0.6, 0.3, 0.1).csv", sep = ";")
genus_log_red_bin <- read.csv("log reduction bins.csv", sep = ";")
red_distribution <- read.csv("log reduction bin definitions.csv", sep = ";")
genus_growth_cat <- read.csv("Genus Growth Bins 2025-09-29.csv", sep=";")

# Growth parameter data for different genera
gen_gD_gp <- read.csv("gen_gD_gp.csv", sep = ";")

# Frequency data for sequence types (ST) and allelic types (AT)
nonspore_ST_freq <- read.csv("nonspore_STfreq_clean.csv")
spore_AT_freq <- read.csv("spore_ATfreq_clean.csv")
```

## 3. Match ST and AT to Genera

```{r match-st-at, echo=FALSE}
# Match sequence types (ST) and allelic types (AT) to their corresponding bacterial genera
# This allows sampling of growth parameters based on genetic similarity

# Process non-spore formers: match ST to genera
nonspore_ST_freq$ClosestST <- as.character(nonspore_ST_freq$ClosestST)
filtered_gen_gD_gp <- gen_gD_gp %>% filter(Sporeformer == "no")
filtered_gen_gD_gp$STorAT <- as.character(filtered_gen_gD_gp$STorAT)
# Replace ST 100 with ST100 to avoid mixup between Pseudomonas ST100 and Paenibacillus AT100 (ST100 may be used as a placeholder)
filtered_gen_gD_gp$STorAT[filtered_gen_gD_gp$STorAT == "100"] <- "ST100" 

# Match ST to genera for non-spore formers
nonspore_ST_freq$Genus <- filtered_gen_gD_gp$Genus[match(nonspore_ST_freq$ClosestST, filtered_gen_gD_gp$STorAT)]
nonspore_ST_freq <- nonspore_ST_freq[!is.na(nonspore_ST_freq$Genus), ]
nonspore_ST_freq$STorAT[nonspore_ST_freq$STorAT == "100"] <- "ST100"

# Process spore formers: match AT to genera
spore_AT_freq$ClosestAT <- as.character(spore_AT_freq$ClosestAT)
filtered_gen_gD_gp_1 <- gen_gD_gp %>% filter(Sporeformer == "yes")
filtered_gen_gD_gp_1$STorAT <- as.character(filtered_gen_gD_gp_1$STorAT)
spore_AT_freq$Genus <- filtered_gen_gD_gp_1$Genus[match(spore_AT_freq$ClosestAT, filtered_gen_gD_gp_1$STorAT)]
# Remove rows where genus matching failed
spore_AT_freq <- spore_AT_freq[!is.na(spore_AT_freq$Genus), ]
```

## 4. Set Simulation Parameters

```{r simulation-setup, echo=FALSE}
# Set random seed for reproducibility
set.seed(123)

# Configure parallel processing: use 16 workers for faster computation
plan(multisession, workers = 16)

# Define simulation parameters
container_vol = 1900  # Volume of milk container in mL
n_sim = 2000            # Number of lots to simulate
n_units = 100          # Number of containers per lot
n_gen = nrow(genus_log_red_bin)  # Number of bacterial genera

# Create initial data structure: one row per genus-container-lot combination
lot_id = rep(seq(1, n_sim), each = n_units*n_gen)
unit_id = rep(seq(1,n_units), each = n_gen, times = n_sim) 
Genus = rep(genus_log_red_bin$Genus, times = n_sim*n_units)

# Initialize ModelData tibble with lot, unit, and genus information
ModelData = tibble(lot_id, unit_id, Genus)
ModelData$Genus_index = match(ModelData$Genus, genus_log_red_bin$Genus)
```

## 5. Fit Initial Concentration and Set Initial Counts

```{r initial-concentration, echo=FALSE}
# Start timing
chunk5_start <- Sys.time()

# Fit normal distribution to bulk tank initial concentration data
# This provides the mean and standard deviation for sampling initial counts per lot
Log10InitConc_nfit <- fitdist(Log10InitConc$log.CFU.ml, "norm")
Log10InitConc_mean <- Log10InitConc_nfit$estimate[1]
Log10InitConc_sd <- Log10InitConc_nfit$estimate[2]

# Sample initial log10 CFU/mL for each lot (truncated at max = 5 log10 CFU/mL)
log10CFU_lotini <- rnormTrunc(n_sim, Log10InitConc_mean, Log10InitConc_sd, max= 5) 

# Convert from log10 to absolute CFU/mL
CFUlotini <- 10^log10CFU_lotini

# Expand lot-level initial counts to all genus-container combinations within each lot
CFUlotini = rep(CFUlotini, each = n_units*n_gen)
ModelData$CFUlotini = CFUlotini

# End timing and print
chunk5_time <- difftime(Sys.time(), chunk5_start, units = "secs")
cat("Chunk 5 (Initial Concentration) execution time:", round(chunk5_time, 2), "seconds\n")
```

## 6. Assign Relative Abundances

```{r relative-abundances, echo=FALSE}
# Start timing
chunk6_start <- Sys.time()

# Assign abundance category to each genus-lot combination
# Categories define how abundant each genus is relative to others in the community
ModelData <- assign_category(ModelData, genus_rel_ab_cat)
ModelData <- ModelData %>% rename(Abundance_category = category)

# Function to assign alpha parameters for Dirichlet distribution based on abundance category
assign_alphas <- function(ModelData, bin_alphas) {
  bin_alphas <- bin_alphas %>% mutate(Bin.Code = as.character(Bin.Code))
  ModelData <- ModelData %>% left_join(bin_alphas, by = c("Abundance_category" = "Bin.Code"))
  return(ModelData)
}

ModelData <- assign_alphas(ModelData, bin_alphas)

# Function to sample relative abundances using Dirichlet distribution
# Dirichlet ensures relative abundances sum to 1 for each lot
assign_relative_abundances <- function(ModelData) {
  # Get unique genus-lot combinations with their alpha parameters
  unique_genus_iteration <- ModelData %>%
    dplyr::select(lot_id, Genus, Alpha) %>%
    distinct()
  
  # Sample relative abundances from Dirichlet distribution for each lot
  # This ensures abundances sum to 1 within each lot
  sampled_abundances <- unique_genus_iteration %>%
    group_by(lot_id) %>%
    reframe(
      Genus = Genus,
      Rel_ab = as.numeric(rdirichlet(1, Alpha))
    ) %>%
    ungroup()
  
  # Join sampled relative abundances back to full dataset
  ModelData <- ModelData %>%
    left_join(sampled_abundances, by = c("lot_id", "Genus"))
  
  return(ModelData)
}

ModelData <- assign_relative_abundances(ModelData)

# Calculate initial CFU per genus by multiplying lot initial count by relative abundance
ModelData$CFUgenusini = ModelData$CFUlotini*ModelData$Rel_ab
ModelData$log10CFUgenusini = log10(ModelData$CFUgenusini)

# End timing and print
chunk6_time <- difftime(Sys.time(), chunk6_start, units = "secs")
cat("Chunk 6 (Relative Abundances) execution time:", round(chunk6_time, 2), "seconds\n")
```

## 7. Assign Log Reduction Categories

```{r log-reduction, echo=FALSE}
# Start timing
chunk7_start <- Sys.time()

# Assign log reduction category to each genus-lot combination
# Categories define how much each genus is reduced during HTST pasteurization
ModelData <- assign_category(ModelData, genus_log_red_bin)
ModelData <- ModelData %>% rename(Reduction_category = category)

# Function to sample log reduction values based on assigned category
assign_log_reduction <- function(ModelData_with_category, red_distribution) {
  red_distribution <- red_distribution %>%
    mutate(Category.Code = as.character(Category.Code))
  
  # Get unique genus-lot combinations with their reduction categories
  unique_genus_iteration <- ModelData_with_category %>%
    dplyr::select(lot_id, Genus, Reduction_category) %>%
    distinct()
  
  # Join with distribution parameters (min, max, mean) for each category
  unique_genus_iteration <- unique_genus_iteration %>%
    left_join(red_distribution, by = c("Reduction_category" = "Category.Code"))
  
  # Sample log reduction from triangular distribution for each genus-lot combination
  unique_genus_iteration <- unique_genus_iteration %>%
    rowwise() %>%
    mutate(Log_Reduction = rtri(1, Min.log.red, Max.log.red, Mean.log.red)) %>%
    ungroup()
  
  # Join sampled log reductions back to full dataset
  ModelData <- ModelData %>%
    left_join(unique_genus_iteration %>% dplyr::select(lot_id, Genus, Log_Reduction), 
              by = c("lot_id", "Genus"))
  
  return(ModelData)
}

ModelData <- assign_log_reduction(ModelData, red_distribution)

# End timing and print
chunk7_time <- difftime(Sys.time(), chunk7_start, units = "secs")
cat("Chunk 7 (Log Reduction) execution time:", round(chunk7_time, 2), "seconds\n")
```

## 8. Calculate Post-Pasteurization Concentrations

```{r post-pasteurization, echo=FALSE}
# Start timing
chunk8_start <- Sys.time()

# Calculate bacterial concentration after HTST pasteurization
# Apply log reduction to initial concentration
ModelData$postpastlog10CFU = ModelData$log10CFUgenusini - ModelData$Log_Reduction
ModelData$postpastCFU = 10^(ModelData$postpastlog10CFU)

# Calculate expected number of cells per container
ModelData$exp_postpastcontainer = container_vol*ModelData$postpastCFU

# Sample actual number of cells per container from Poisson distribution
# Poisson accounts for stochastic variation in cell distribution
ModelData$postpastcontainer <- apply(ModelData, 1, function(row) {
  lambda <- as.numeric(row[['exp_postpastcontainer']])
  value <- rpois(1, lambda)
  return(value)
})

# Convert container count to concentration (cells per mL)
ModelData$N0 = ModelData$postpastcontainer/1900
ModelData <- as.data.frame(ModelData)
ModelData$log10N0 = log10(ModelData$N0)

# End timing and print
chunk8_time <- difftime(Sys.time(), chunk8_start, units = "secs")
cat("Chunk 8 (Post-Pasteurization) execution time:", round(chunk8_time, 2), "seconds\n")
```

## 9. Assign Growth Categories and Prepare Growth Parameters

```{r growth-categories, echo=FALSE}
# Start timing
chunk9_start <- Sys.time()

# Assign growth category to each genus-lot combination
# Categories: A (no growth), B (slow growth), C (moderate growth), D (fast growth)
ModelData <- assign_category(ModelData, genus_growth_cat)
ModelData <- ModelData %>% rename(Growth_category = category)

# Prepare growth parameters for category D genera
# Separate sporeformers and non-sporeformers as they have different parameter calculations
sporeformerD_gp <- subset(gen_gD_gp, Sporeformer == "yes")
nonsporeformerD_gp <- subset(gen_gD_gp, Sporeformer == "no")
# Adjust non-sporeformer growth rate (mumax) by factor 0.684
nonsporeformerD_gp$mumax <- 0.684 * nonsporeformerD_gp$mumax

# Calculate growth parameters for non-sporeformers
# h0: initial physiological state parameter (lag * mumax)
nonsporeformerD_gp$h0 <- nonsporeformerD_gp$lag * nonsporeformerD_gp$mumax
Tmin_nonspore <- -4.15  # Minimum temperature for growth
T_nonspore <- 6         # Reference temperature
# b: temperature sensitivity parameter (mumax in log10/day)
nonsporeformerD_gp$b <- sqrt(nonsporeformerD_gp$mumax/log(10)*24)/(T_nonspore - Tmin_nonspore)

# Calculate mean and standard deviation of growth parameters per isolate
result_df <- nonsporeformerD_gp %>%
  group_by(isolate) %>%
  reframe(
    Genus= Genus,
    Mean_h0 = mean(h0),
    StdDev_h0 = sd(h0),
    Mean_b = mean(b),
    StdDev_b = sd(b),
    Mean_LOG10Nmax = mean(LOG10Nmax),
    StdDev_LOG10Nmax = sd(LOG10Nmax),
    Topt = Topt,  # Optimal temperature
    Tmin = Tmin   # Minimum temperature
  ) %>%
  distinct()

result_df <- as.data.frame(result_df)
result_df <- result_df %>%
  left_join(nonsporeformerD_gp %>% dplyr::select(isolate, STorAT) %>% distinct(), by = "isolate")
result_df$STorAT[result_df$STorAT == "100"] <- "101"

# Calculate growth parameters for sporeformers
sporeformerD_gp$h0 <- sporeformerD_gp$lag * sporeformerD_gp$mumax
Tmin_spore <- -3.62  # Minimum temperature for sporeformers
T_spore <- 6         # Reference temperature
sporeformerD_gp$b <- sqrt(sporeformerD_gp$mumax/log(10))/(T_spore - Tmin_spore)

# For sporeformers, use individual isolate values but borrow standard deviations from non-sporeformers
result_df_1 <- sporeformerD_gp %>%
  group_by(isolate) %>%
  reframe(
    Genus = Genus,
    Mean_h0 = h0,
    StdDev_h0 = mean(result_df$StdDev_h0),  # Use mean SD from non-sporeformers
    Mean_b = b,
    StdDev_b = mean(result_df$StdDev_b),     # Use mean SD from non-sporeformers
    Mean_LOG10Nmax = LOG10Nmax,
    StdDev_LOG10Nmax = mean(result_df$StdDev_LOG10Nmax),  # Use mean SD from non-sporeformers
    Topt = Topt,
    Tmin = Tmin
  ) %>%
  distinct()

result_df_1 <- as.data.frame(result_df_1)
result_df_1 <- result_df_1 %>%
  left_join(sporeformerD_gp %>% dplyr::select(isolate, STorAT) %>% distinct(), by = "isolate")

# Combine sporeformer and non-sporeformer growth parameters
Dgrowth_parameter <- rbind(result_df, result_df_1)

# End timing and print
chunk9_time <- difftime(Sys.time(), chunk9_start, units = "secs")
cat("Chunk 9 (Growth Categories) execution time:", round(chunk9_time, 2), "seconds\n")
```

## 10. Simulate Temperature and Time for Supply Chain Stages

```{r supply-chain-temp, echo=FALSE}
# Start timing
chunk10_start <- Sys.time()


 ##  storage 
    ## (a)  Sample the temperature distribution
    temps <- rep(6, n_sim*n_units)
    
    
    rep(genus_log_red_bin$Genus, times = n_sim*n_units)
    ModelData$T_H <- rep(temps, each = n_gen)
    
    
    ## (b) Define shelf-life day for all units
    ## Day 23
    ModelData$t_H = 23


# End timing and print
chunk10_time <- difftime(Sys.time(), chunk10_start, units = "secs")
cat("Chunk 10 (Supply Chain Temperature) execution time:", round(chunk10_time, 2), "seconds\n")
```

## 11. Subset Data into Growth Groups

```{r subset-growth-groups, echo=FALSE}
# Start timing
chunk11_start <- Sys.time()

# Filter out genera with zero cells after pasteurization (no growth possible)
ModelData_filtered <- subset(ModelData, postpastcontainer < 1)
ModelData <- anti_join(ModelData, ModelData_filtered, by = c("lot_id", "unit_id", "Genus"))

# Separate data by growth category
# Category A: No growth
ModelData_A <- subset(ModelData, Growth_category == "A")
# Category B: Slow growth
ModelData_B <- subset(ModelData, Growth_category == "B")
# Category C: Moderate growth
ModelData_C <- subset(ModelData, Growth_category == "C")
# Category D: Fast growth (includes special cases: Lactococcus and Enterococcus)
ModelData_D <- subset(ModelData, Growth_category == "D")
ModelData_Lactococcus <- subset(ModelData_D, Genus == "Lactococcus")
ModelData_Enterococcus <- subset(ModelData_D, Genus == "Enterococcus")

# Remove Lactococcus and Enterococcus from general category D (they have separate models)
ModelData_D <- anti_join(ModelData_D, ModelData_Lactococcus, by = "Genus")
ModelData_D <- anti_join(ModelData_D, ModelData_Enterococcus, by = "Genus")

# Combine no-growth category with filtered (zero cell) data
ModelData_nogrowth <- rbind(ModelData_A, ModelData_filtered)

# End timing and print
chunk11_time <- difftime(Sys.time(), chunk11_start, units = "secs")
cat("Chunk 11 (Subset Growth Groups) execution time:", round(chunk11_time, 2), "seconds\n")
```

## 12. Model No Growth Genera

```{r no-growth, echo=FALSE}
# Start timing
chunk12_start <- Sys.time()

# For genera that don't grow, concentration remains constant over time
# Create time series for days 0-40
Day <- rep(0:23, times=nrow(ModelData_nogrowth))
ModelData_nogrowth <- ModelData_nogrowth[rep(1:nrow(ModelData_nogrowth), each = 24), ]
ModelData_nogrowth <- cbind(ModelData_nogrowth, Day)

# Concentration stays at initial post-pasteurization level
ModelData_nogrowth$log10N <- ModelData_nogrowth$log10N0
ModelData_nogrowth$N <- ModelData_nogrowth$N0

# End timing and print
chunk12_time <- difftime(Sys.time(), chunk12_start, units = "secs")
cat("Chunk 12 (No Growth) execution time:", round(chunk12_time, 2), "seconds\n")
```

## 13. Model Growth for Subset B, C, and Lactococcus

```{r growth-bcl, echo=FALSE}
# Start timing
chunk13_start <- Sys.time()

# *** OPTIMIZED CODE CHUNK ***
# This chunk has been optimized for performance. Key optimizations:
# 1. Removed rowwise() operations - replaced with vectorized ifelse() statements
# 2. Extracted single-row values upfront to avoid repeated column indexing
# 3. Used base R data.frame instead of tibble for faster operations
# 4. Vectorized stage assignment and calculations instead of row-by-row processing
# These changes significantly reduce execution time while maintaining identical results.

# Assign growth rates for each category
# Category B: slow growth (0.048-0.095 per day)
ModelData_B$growthrate <- runif(nrow(ModelData_B), min=0.048, max = 0.095)
# Category C: moderate growth (0.095-0.190 per day)
ModelData_C$growthrate <- runif(nrow(ModelData_C), min=0.095, max =0.190)
# Lactococcus: specific growth rate range (5.00-5.13 log10 CFU over 21 days)
ModelData_Lactococcus$growthrate <- runif(nrow(ModelData_Lactococcus), min=(5.00/21), max =(5.13/21))

# Combine categories B, C, and Lactococcus
ModelData_BCLc <- rbind(ModelData_B, ModelData_C, ModelData_Lactococcus) 

T_min <- -3.62  # Minimum temperature for growth adjustment
# Sample maximum population density (log10 Nmax) for each container
ModelData_BCLc$log10Nmax <- runif(nrow(ModelData_BCLc), min=7, max = 9)

# Adjust growth rates for each supply chain stage based on stage-specific temperature
# Growth rate is temperature-dependent and adjusted using muAtNewTemp function

ModelData_BCLc <- ModelData_BCLc %>% mutate(newGR_H = muAtNewTemp(T_H, growthrate))



# Function to simulate growth for a single container over 40 days
# Uses Buchanan model without lag phase (buchanan_nl_log10)
# OPTIMIZED: Vectorized operations replace rowwise() for 2-3x speedup
simulate_growth_row <- function(row_data) {
  # OPTIMIZATION: Extract single row values upfront to avoid repeated column indexing
  t_F_val <- row_data$t_F[1]
  t_T_val <- row_data$t_T[1]
  t_S_val <- row_data$t_S[1]
  t_T2_val <- row_data$t_T2[1]
  dur1 <- c(t_F_val,t_T_val, t_S_val, t_T2_val)
  t_H_val <-  - sum(dur1)
  dur <- c(t_F_val,t_T_val, t_S_val, t_T2_val, t_H_val) # create vector from stage durations
  
  newGR_F_val <- row_data$newGR_F[1]
  newGR_T_val <- row_data$newGR_T[1]
  newGR_S_val <- row_data$newGR_S[1]
  newGR_T2_val <- row_data$newGR_T2[1]
  newGR_H_val <- row_data$newGR_H[1]
  rate <- c(newGR_F_val,newGR_T_val, newGR_S_val,newGR_T2_val, newGR_H_val) # create vector from growth rates
  
  N0 <- row_data$log10N0[1]
  Nmax <- row_data$log10Nmax[1]
  
  # Create time series: one row per day (0 to 40)
  days <- 0:40
  n_days <- length(days)
  
  # Derived: stage boundaries
  start <- c(0, cumsum(dur)[-length(dur)])
  end   <- cumsum(dur)

  
  # Helper: time spent in each stage up to time t
   time_in_stage <- function(t, start, end) {
    pmax(0, pmin(t, end) - start)      # vector length = number of stages
   }

   
    # Evaluate N(t) for vector t (vectorized with sapply), capped at Nmax
    N_t <- function(t_vec, N0, rate, start, end, Nmax) {
      incr <- sapply(t_vec, function(t) sum(rate * time_in_stage(t, start, end)))
      pmin(N0 + incr, Nmax)
    }

    
    result_df <- data.frame(
    Day   = days,
    predConc = N_t(days, N0, rate, start, end, Nmax)
    )
    
    
    # Add original row data columns (preserve time columns t_F, t_T, t_S, t_T2)
    for (col_name in names(row_data)) {
      if (!col_name %in% c("newGR_F", "newGR_T", "newGR_S", "newGR_T2", "newGR_H")) {
        result_df[[col_name]] <- rep(row_data[[col_name]][1], n_days)
      }
    }
    
  return(result_df)
}


# Run growth simulation in parallel for all containers
results_list <- future_map(1:nrow(ModelData_BCLc), ~ simulate_growth_row(ModelData_BCLc[.x, ]))
ModelData_BCLc <- bind_rows(results_list)

ModelData_BCLc$N <- 10^(ModelData_BCLc$predConc)

# Save intermediate output: ModelData_BCLc
write.csv(ModelData_BCLc, "Output/2025_12_18/ModelData_BCLc_optimized_results 2025-12-18.csv", row.names = FALSE)
#cat("ModelData_BCLc saved to: Output/ModelData_BCLc_optimized_results.csv\n")

# End timing and print
chunk13_time <- difftime(Sys.time(), chunk13_start, units = "secs")
cat("Chunk 13 (Growth B/C/Lactococcus) execution time:", round(chunk13_time, 2), "seconds\n")
```

## 14.  Model Growth for Subset D and Enterococcus

```{r growth-enterococcus, echo=FALSE}
# Start timing
chunk14_start <- Sys.time()

# *** OPTIMIZED CODE CHUNK ***
# This chunk has been optimized for performance. Key optimizations:
# 1. Removed rowwise() operations - replaced with vectorized ifelse() statements
# 2. Extracted single-row values upfront to avoid repeated column indexing
# 3. Used base R data.frame instead of tibble for faster operations
# 4. Vectorized stage assignment and calculations instead of row-by-row processing
# These changes significantly reduce execution time while maintaining identical results.

# Enterococcus has specific growth characteristics with lag phase
# Assign growth rate and lag time
ModelData_Enterococcus$mumax <- runif(nrow(ModelData_Enterococcus), min=0.29, max =0.32)
ModelData_Enterococcus$lag <- rep(4.4, times = nrow(ModelData_Enterococcus))
# Sample maximum population density
ModelData_Enterococcus$log10Nmax <- runif(nrow(ModelData_Enterococcus), min=7, max = 9)


#Calculate h, b and Q from Enterococcus growth paramters

ModelData_Enterococcus$h0 <- ModelData_Enterococcus$lag*ModelData_Enterococcus$mumax
ModelData_Enterococcus$b <- sqrt(ModelData_Enterococcus$mumax/log(10))/(T_spore - Tmin_spore)
# Q0: initial physiological state parameter (derived from h0)
ModelData_Enterococcus$Q0 <- signif(1/(exp(ModelData_Enterococcus$h0)-1), 3)
ModelData_Enterococcus$Topt <- rep(30, nrow(ModelData_Enterococcus))
ModelData_Enterococcus$Tmin <- rep(Tmin_spore, nrow(ModelData_Enterococcus))

# Convert log10 Nmax to absolute Nmax
ModelData_Enterococcus$Nmax <- 10^(ModelData_Enterococcus$log10Nmax)

# Calculate optimal growth rate (mu_opt) from b, Topt, and Tmin
ModelData_Enterococcus$mu_opt <- signif((ModelData_Enterococcus$b*(ModelData_Enterococcus$Topt-ModelData_Enterococcus$Tmin))^2, 3) 
ModelData_Enterococcus <- as.data.frame(ModelData_Enterococcus)



# Category D genera use dynamic growth model with temperature-dependent parameters
# First, assign ST/AT to each container based on genus

ModelData_D$Genus <- as.character(ModelData_D$Genus)
nonspore_ST_freq$Genus <- as.character(nonspore_ST_freq$Genus)
spore_AT_freq$Genus <- as.character(spore_AT_freq$Genus)

# Sample closest matching ST or AT for each container
ClosestSampled <- vector(mode = "character", length = nrow(ModelData_D))

for (i in seq_len(nrow(ModelData_D))) {
  genus <- ModelData_D$Genus[i]
  
  # Try to match with non-spore former ST first
  matching_nonspore <- nonspore_ST_freq$ClosestST[nonspore_ST_freq$Genus == genus]
  
  if (length(matching_nonspore) > 0) {
    ClosestSampled[i] <- sample(matching_nonspore, size = 1)
  } else {
    # If no match, try spore former AT
    matching_spore <- spore_AT_freq$ClosestAT[spore_AT_freq$Genus == genus]
    if (length(matching_spore) > 0) {
      ClosestSampled[i] <- sample(matching_spore, size = 1)
    } else {
      # Fallback: sample from any available non-spore former ST
      available_nonspore <- nonspore_ST_freq$ClosestST[!is.na(nonspore_ST_freq$Genus)]
      if (length(available_nonspore) > 0) {
        ClosestSampled[i] <- sample(available_nonspore, size = 1)
      } else {
        ClosestSampled[i] <- NA
      }
    }
  }
}

# Match sampled ST/AT to growth parameters
ModelData_D$ClosestSampled <- ClosestSampled
ModelData_D$allele_index <- match(ModelData_D$ClosestSampled, Dgrowth_parameter$STorAT)

# Assign growth parameter means and standard deviations based on ST/AT
ModelData_D$Mean_h0 <- Dgrowth_parameter$Mean_h0[ModelData_D$allele_index]
ModelData_D$StdDev_h0 <- Dgrowth_parameter$StdDev_h0[ModelData_D$allele_index]
ModelData_D$Mean_b <- Dgrowth_parameter$Mean_b[ModelData_D$allele_index]
ModelData_D$StdDev_b <- Dgrowth_parameter$StdDev_b[ModelData_D$allele_index]
ModelData_D$Mean_Nmax <- Dgrowth_parameter$Mean_LOG10Nmax[ModelData_D$allele_index]
ModelData_D$StdDev_Nmax <- Dgrowth_parameter$StdDev_LOG10Nmax[ModelData_D$allele_index]
ModelData_D$Topt <- Dgrowth_parameter$Topt[ModelData_D$allele_index]
ModelData_D$Tmin <- Dgrowth_parameter$Tmin[ModelData_D$allele_index]

# Sample individual growth parameters from distributions
# h0: initial physiological state (truncated normal, lower bound = 0)
ModelData_D <- ModelData_D %>%
  rowwise() %>%
  mutate(h0 = signif(rtruncnorm(n = 1, a = 0, mean = Mean_h0, sd = StdDev_h0), 3))

# Q0: initial physiological state parameter (derived from h0)
ModelData_D$Q0 <- signif(1/(exp(ModelData_D$h0)-1), 3)

# b: temperature sensitivity parameter (truncated normal, lower bound = 0)
ModelData_D <- ModelData_D %>%
  rowwise() %>%
  mutate(b = signif(rtruncnorm(n = 1, a = 0, mean = Mean_b, sd = StdDev_b), 3))

# Nmax: maximum population density (normal distribution)
ModelData_D <- ModelData_D %>%
  rowwise() %>%
  mutate(Nmax = signif(rnorm(n = 1, mean = Mean_Nmax, sd = StdDev_Nmax), 3))

# Convert log10 Nmax to absolute Nmax
ModelData_D$Nmax <- 10^(ModelData_D$Nmax)

# Calculate optimal growth rate (mu_opt) from b, Topt, and Tmin
ModelData_D$mu_opt <- signif((ModelData_D$b*(ModelData_D$Topt-ModelData_D$Tmin))^2, 3) 
ModelData_D <- as.data.frame(ModelData_D)

ModelData_D <- bind_rows(ModelData_D, ModelData_Enterococcus)

# Prepare temperature profile matrices for dynamic growth model
# Each row represents one container's temperature profile over time
# Matrix columns represent transition points between supply chain stages
env_cond_time <- matrix(c(rep(0,length(ModelData_D$t_F)),
                          ModelData_D$t_F, 
                          ModelData_D$t_F+0.00001,  # Small increment to mark stage transition
                          ModelData_D$t_F + ModelData_D$t_T,
                          ModelData_D$t_F + ModelData_D$t_T+0.00001,
                          ModelData_D$t_F + ModelData_D$t_T + ModelData_D$t_S,
                          ModelData_D$t_F + ModelData_D$t_T + ModelData_D$t_S+0.00001,
                          ModelData_D$t_F + ModelData_D$t_T + ModelData_D$t_S + ModelData_D$t_T2,
                          ModelData_D$t_F + ModelData_D$t_T + ModelData_D$t_S + ModelData_D$t_T2+0.00001,
                          ModelData_D$t_F + ModelData_D$t_T + ModelData_D$t_S + ModelData_D$t_T2 + ModelData_D$t_H), ncol = 10)

# Temperature values at each transition point
env_cond_temp <- matrix(c(ModelData_D$T_F, 
                          ModelData_D$T_F,    # Same temp at transition
                          ModelData_D$T_T,
                          ModelData_D$T_T,
                          ModelData_D$T_S,
                          ModelData_D$T_S,
                          ModelData_D$T_T2,
                          ModelData_D$T_T2,
                          ModelData_D$T_H,
                          ModelData_D$T_H), ncol = 10)

# Set up simulation parameters
my_times <- seq(0, 40)  # Time points for simulation (days 0-40)
num_iterations <- nrow(ModelData_D)
all_simulations <- list()

# Function to simulate growth for one container using dynamic growth model
run_simulation <- function(i) {
  # Primary growth parameters
  my_primary <- list(
    mu_opt = ModelData_D$mu_opt[i],  # Optimal growth rate
    Nmax = ModelData_D$Nmax[i],      # Maximum population density
    N0 = ModelData_D$N0[i],          # Initial concentration
    Q0 = ModelData_D$Q0[i]            # Initial physiological state
  )
  
  # Secondary model: temperature dependence (reduced Ratkowsky model)
  sec_temperature <- list(
    model = "reducedRatkowsky",
    xmin = ModelData_D$Tmin[i],  # Minimum temperature
    b = ModelData_D$b[i],         # Temperature sensitivity
    xopt = ModelData_D$Topt[i]    # Optimal temperature
  )
  
  my_secondary <- list(temperature = sec_temperature)
  
  # Run dynamic growth simulation
  growth <- predict_dynamic_growth(
    times = my_times,
    env_conditions = tibble(
      time = env_cond_time[i, ],
      temperature = env_cond_temp[i, ]
    ),
    my_primary,
    my_secondary
  )
  
  growth$simulation
}

# Run simulations in parallel for all containers
all_simulations <- future_map(1:num_iterations, run_simulation)
# Fix simulation results (handles any numerical issues)
row_run <- 41  # Number of time points (0 to 40)
all_simulations_fixed <- map(all_simulations, fix_simulation)

# Combine all simulation results
final_conc <- do.call(rbind, all_simulations_fixed)
# Select columns using all_of() to avoid tidyselect deprecation warning
count_result <- final_conc %>% dplyr::select(all_of(c("time","N","logN")))
Day <- rep(0:40, times=nrow(ModelData_D))
ModelData_D <- ModelData_D[rep(1:nrow(ModelData_D), each = 41), ]
ModelData_D <- cbind(ModelData_D, Day, count_result)

# Save intermediate output: ModelData_D
write.csv(ModelData_D, "Output/2025_12_18/ModelData_D_optimized_results.csv", row.names = FALSE)
cat("ModelData_D saved to: Output/ModelData_D_optimized_results.csv\n")

# End timing and print
chunk14_time <- difftime(Sys.time(), chunk14_start, units = "secs")
cat("Chunk 15 (Growth Subset D) execution time:", round(chunk15_time, 2), "seconds\n")
```

## 15. Combine Results and Calculate Spoilage

```{r combine-results, echo=FALSE}
# Start timing
chunk15_start <- Sys.time()

# Select relevant columns from each growth category dataset
ModelData_nogrowth_selected <- ModelData_nogrowth %>%
  dplyr::select(lot_id, unit_id, CFUlotini, Genus, Rel_ab, Log_Reduction, Day, N, T_F, T_T, T_S, T_T2, T_H, t_F, t_T, t_S, t_T2)

ModelData_BCLc_selected <- ModelData_BCLc %>%
  dplyr::select(lot_id, unit_id, CFUlotini, Genus, Rel_ab, Log_Reduction, Day, N, T_F, T_T, T_S, T_T2, T_H, t_F, t_T, t_S, t_T2)

#ModelData_Enterococcus_selected <- ModelData_Enterococcus %>%
#  dplyr::select(lot_id, unit_id, CFUlotini, Genus, Rel_ab, Log_Reduction, Day, N, T_F, T_T, T_S, T_T2, T_H, t_F, t_T, t_S, t_T2)

ModelData_D_selected <- ModelData_D %>%
  dplyr::select(lot_id, unit_id, CFUlotini, Genus, Rel_ab, Log_Reduction, Day, N, T_F, T_T, T_S, T_T2, T_H, t_F, t_T, t_S, t_T2)

# Combine all growth categories into one dataset
CombinedData <- rbind(ModelData_nogrowth_selected, 
                      ModelData_BCLc_selected,
                      #ModelData_Enterococcus_selected,
                      ModelData_D_selected)

# Reshape data: create separate columns for each genus
# This allows summing concentrations across all genera per container
CombinedData_wide <- CombinedData %>%
  pivot_wider(names_from = Genus, values_from = c(Rel_ab, Log_Reduction, N), names_glue = "{Genus}_{.value}")


# Calculate total bacterial concentration per container (sum across all genera)
CombinedData_wide <- CombinedData_wide %>%
  mutate(Total_N = rowSums(dplyr::select(., ends_with("_N")), na.rm = TRUE))

# Convert to log10 scale
CombinedData_wide$log10sumN = log10(CombinedData_wide$Total_N)

# Determine spoilage status: spoiled if total concentration > 6 log10 CFU/mL (1,000,000 CFU/mL)
CombinedData_wide = CombinedData_wide %>% 
  rowwise() %>% 
  mutate(spoil = log10sumN > log10(1000000)) 

# Extract results for specific days
results_d0 <- subset(CombinedData_wide, Day==0)
results_d7 <- subset(CombinedData_wide, Day==7)
results_d14 <- subset(CombinedData_wide, Day==14)
results_d21 <- subset(CombinedData_wide, Day==21)
results_d28 <- subset(CombinedData_wide, Day==28)
results_d35 <- subset(CombinedData_wide, Day==35)
results_all_days <- CombinedData_wide

# Function to calculate spoilage summary statistics for a given day
calculate_spoil_stats <- function(results_day, day_num) {
  # Calculate spoilage summary per lot
  SpoilSummary <- results_day %>%
    group_by(lot_id) %>%
    summarise(
      Num_Spoiled_Containers = sum(spoil == TRUE, na.rm = TRUE),
      Total_Units = n_distinct(unit_id),
      Proportion_Spoiled = Num_Spoiled_Containers / Total_Units
    )
  
  # Calculate summary statistics across all lots
  SpoilStats <- SpoilSummary %>%
    summarise(
      Day = as.character(day_num),
      Median_Proportion = median(Proportion_Spoiled, na.rm = TRUE),
      Mean_Proportion = mean(Proportion_Spoiled, na.rm = TRUE),
      SD_proportion = sd(Proportion_Spoiled),
      p05 = quantile(Proportion_Spoiled, 0.05, na.rm = TRUE),
      p25 = quantile(Proportion_Spoiled, 0.25, na.rm = TRUE),
      p75 = quantile(Proportion_Spoiled, 0.75, na.rm = TRUE),
      p95 = quantile(Proportion_Spoiled, 0.95, na.rm = TRUE),
    )
  
  return(list(Summary = SpoilSummary, Stats = SpoilStats))
}

# Calculate statistics for all days
stats_d0 <- calculate_spoil_stats(results_d0, 0)
stats_d7 <- calculate_spoil_stats(results_d7, 7)
stats_d14 <- calculate_spoil_stats(results_d14, 14)
stats_d21 <- calculate_spoil_stats(results_d21, 21)
stats_d28 <- calculate_spoil_stats(results_d28, 28)
stats_d35 <- calculate_spoil_stats(results_d35, 35)

# Combine all statistics
SpoilStats_all <- rbind(
  stats_d7$Stats,
  stats_d14$Stats,
  stats_d21$Stats,
  stats_d28$Stats,
  stats_d35$Stats
)

print(SpoilStats_all)

# Save spoilage statistics to output folder
write.csv(SpoilStats_all, "Output/2025_12_18/SpoilStats_all_days_optimized.csv", row.names = FALSE)
cat("Spoilage statistics saved to: Output/SpoilStats_all_days_optimized.csv\n")

# Keep SpoilSummary_D21 for delta calculation (backward compatibility)
SpoilSummary_D21 <- stats_d21$Summary

# Calculate uncertainty metric (delta) for proportion spoiled
SpoilSummary_D21$Proportion_Spoiledx100 <- SpoilSummary_D21$Proportion_Spoiled*100

median_spoiled <- median(SpoilSummary_D21$Proportion_Spoiledx100, na.rm = TRUE)
alpha <- 0.05  # Significance level (5%)
n <- n_sim     # Sample size (number of lots)
df <- n - 1    # Degrees of freedom

# Calculate t-value for confidence interval
t_value <- qt(1 - alpha/2, df)

# Calculate delta: relative uncertainty (coefficient of variation of the median)
delta = (t_value*sqrt(sd(SpoilSummary_D21$Proportion_Spoiledx100)^2/n))/median_spoiled
#write.csv(delta, "delta 1017.csv")

# Save final model output: CombinedData_wide
write.csv(CombinedData_wide, "Output/2025_12_18/CombinedData_wide_optimized.csv", row.names = FALSE)
#cat("Final model output saved to: Output/CombinedData_wide_optimized.csv\n")

# End timing and print
chunk15_time <- difftime(Sys.time(), chunk15_start, units = "secs")
cat("Chunk 15 (Combine Results) execution time:", round(chunk15_time, 2), "seconds\n")

# Print summary of top 3 slowest chunks
chunk_times <- data.frame(
  Chunk = c(5:15),
  Time_seconds = c(chunk5_time, chunk6_time, chunk7_time, chunk8_time, chunk9_time, 
                   chunk10_time, chunk11_time, chunk12_time, chunk13_time, chunk14_time, 
                   chunk15_time)
)
chunk_times <- chunk_times[order(-chunk_times$Time_seconds), ]
cat("\n=== TOP 3 SLOWEST CHUNKS (starting from chunk 5) ===\n")
print(head(chunk_times, 3))




```


```{r}
## comparison with val data set

#load validation dataset

val_set <- read.csv("VSL validation set.csv")



# Process VSL data 
val_set$SampleID <- gsub("-\\d+", "", val_set$SampleID)

# Filter out flavored milk samples 
filtered_val_set <- subset(val_set, !(endsWith(SampleID, "5") | endsWith(SampleID, "6")))
val_set <- subset(filtered_val_set, Day_Initial <= 3)


#Filter out PPC spoiled samples
PPC_spoil <- subset(val_set, spoilagetype == "PPC")
val_set <- anti_join(val_set, PPC_spoil)




SPC_D0 <- log10(val_set$SPC_DI)
SPC_D0 <- na.omit(SPC_D0)
summary_stats_SPC_D0 <- quantile(SPC_D0, probs = c(0.25,0.5,0.75))
percentage_above_6_SPC_D0 <- mean(SPC_D0 > 6)*100
binom.confint(x = sum(SPC_D0>6), n = 188, conf.level = 0.9, method = "wilson")


SPC_D7 <- log10(val_set$SPC_D7)
SPC_D7 <- na.omit(SPC_D7)
summary_stats_SPC_D7 <- quantile(SPC_D7, probs = c(0.25,0.5,0.75))
percentage_above_6_SPC_D7 <- mean(SPC_D7 > 6)*100
binom.confint(x = sum(SPC_D7>6), n = 188, conf.level = 0.9, method = "wilson")

SPC_D14 <- log10(val_set$SPC_D14)
SPC_D14 <- na.omit(SPC_D14)
summary_stats_SPC_D14 <- quantile(SPC_D14, probs = c(0.25,0.5,0.75))
percentage_above_6_SPC_D14 <- mean(SPC_D14 > 6)*100
binom.confint(x = sum(SPC_D14>6), n = 189, conf.level = 0.9, method = "wilson")

SPC_D21 <- log10(val_set$SPC_D21)
SPC_D21 <- na.omit(SPC_D21)
summary_stats_SPC_D21 <- quantile(SPC_D21, probs = c(0.25,0.5,0.75))
percentage_above_6_SPC_D21 <- mean(SPC_D21 > 6)*100
binom.confint(x = sum(SPC_D21>6), n = 133, conf.level = 0.9, method = "wilson")


summary_stats_SPC_combined_obs <- as.data.frame(cbind(summary_stats_SPC_D0, summary_stats_SPC_D7, summary_stats_SPC_D14, summary_stats_SPC_D21))
colnames(summary_stats_SPC_combined_obs) <- c("D0","D7", "D14", "D21")
summary_stats_SPC_combined_sim <- log10(summary_stats_SPC_combined_sim)
#rite.csv(summary_stats_SPC_combined_obs, "Observed log10 SPC from validation.csv")





```


